scTOP: physics-inspired order parameters for cellular identification and visualization

### 2023-07-21 manuscript/scTOP (Company of Biologists LaTeX template)/scTOP.tex
added revised manuscript

	\documentclass[aps,superscriptaddress, notitlepage,longbibliography]{revtex4-1}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{physics}
\usepackage[pdftex]{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\setlength {\marginparwidth }{2cm}
\hypersetup{
    colorlinks,
    citecolor=blue,
    filecolor=black,
    linkcolor=red,
    urlcolor=blue
}

\newcommand{\comment}[1]{{\color{blue}#1}}
\newcommand{\edit}[1]{{\color{purple}#1}}
\newcommand{\del}[1]{{\color{red}\st{#1}}}
\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }
% Standardize command font styles and environments
\newcommand{\doccmd}[1]{\texttt{\textbackslash#1}}% command name -- adds backslash automatically
\newcommand{\docopt}[1]{\ensuremath{\langle}\textrm{\textit{#1}}\ensuremath{\rangle}}% optional command argument
\newcommand{\docarg}[1]{\textrm{\textit{#1}}}% (required) command argument
\newcommand{\docenv}[1]{\textsf{#1}}% environment name
\newcommand{\docpkg}[1]{\texttt{#1}}% package name
\newcommand{\doccls}[1]{\texttt{#1}}% document class name
\newcommand{\docclsopt}[1]{\texttt{#1}}% document class option name
\newenvironment{docspec}{\begin{quote}\noindent}{\end{quote}}% command specification environment

@@

\def\be{\begin{equation}} 
\def\ee{\end{equation}}
\newcommand \bea {\begin{eqnarray}} 
\newcommand \eea {\end{eqnarray}} 
\newcommand \sign {\hbox{sign}} 
\newcommand{\nn} {\nonumber}
\newcommand{\mbbE} {\mathbb{E}}
\begin{document}

\title{ scTOP: physics-inspired order parameters for cellular identification and visualization}

\author{Maria Yampolskaya}
\email{mariay@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\author{Michael Herriges}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
\author{Laertis Ikonomou}
\affiliation{Department of Oral Biology. University at Buffalo, The State University of New York, Buffalo, NY, USA}
\affiliation{Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University at Buffalo, The State University of New York, Buffalo, NY, USA}
\author{Darrell Kotton}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
\author{Pankaj Mehta}
\email{pankajm@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{Faculty of Computing and Data Science, Boston University, Boston, MA 02215, USA}
\affiliation{Biological Design Center, Boston University, Boston, MA 02215, USA}

\begin{abstract}
Advances in single-cell RNA-sequencing (scRNA-seq) provide an unprecedented window into cellular identity. The increasing abundance of data requires new theoretical and computational frameworks for understanding cell fate determination, accurately classifying cell fates from expression data, and integrating knowledge from cell atlases. Here, we present single-cell Type Order Parameters (scTOP): a statistical-physics-inspired approach for constructing ``order parameters'' for cell fate given a reference basis of cell types. scTOP can quickly and accurately classify cells at a single-cell resolution, generate interpretable visualizations of developmental trajectories, and assess the fidelity of engineered cells. Importantly, scTOP does this without using feature selection, statistical fitting, or dimensional reduction (e.g., UMAP, PCA, etc.). We illustrate the power of scTOP utilizing a wide variety of human and mouse datasets (both {\it in vivo} and {\it in vitro}). By reanalyzing mouse lung alveolar development data, we characterize a transient perinatal hybrid alveolar type 1/alveolar type 2 (AT1/AT2) cell population that disappears by 15 days post-birth and show that it is transcriptionally distinct from previously identified adult AT2-to-AT1 transitional cell types. Visualizations of lineage tracing data on hematopoiesis using scTOP confirm that a single clone can give rise to as many as three distinct differentiated cell types. We also show how scTOP can quantitatively assess the transcriptional similarity between endogenous and transplanted cells in the context of murine pulmonary cell transplantation. Finally, we provide an easy-to-use Python implementation of scTOP. Our results suggest that physics-inspired order parameters can be an important tool for understanding development and characterizing engineered cells.
\end{abstract}

@@

\section{Introduction}
Humans and other mammals have many specialized cells that arise through a complex developmental process  \cite{zeng_what_2022}. At the blastocyst/epiblast stages of embroynic development, cells are pluripotent and possess the flexibility to develop into fates from all three germ layers \cite{Wolpert_2019}. However, as embryos mature, cellular identity becomes less plastic and more specified through a complex spatiotemporal process involving signaling and other environmental cues. The questions of how and why cells end up in one fate or another are fundamental to developmental biology \cite{stephenson_intercellular_2012, zhu2020principles, rand2021geometry}. More practically, understanding the process of cell fate specification is crucial for developing directed differentiation protocols for engineering cells \emph{in vitro}, with important clinical applications for human health and disease \cite{rossant_impact_2011, rowe2019induced, basil2020cellular}.

One powerful set of techniques for characterizing cellular phenotype is transcriptome-wide measurement using single-cell RNA sequencing (scRNA-seq). Typical scRNA-seq experiments involve thousands of genes and dozens to millions of cells. For this reason, this data is intrinsically high-dimensional and requires specialized computational tools for analysis and interpretation \cite{zhu2020principles, setty2019characterization, schiebinger2019optimal}. This task is especially difficult because the technical challenges of sequencing RNA from individual cells result in data that is noisy and sparse \cite{hicks2018missing, lahnemann_eleven_2020, argelaguet2021computational}. Genes with relatively low expression are often recorded as not being expressed at all, resulting in zero counts called dropouts. Experiment-to-experiment differences in sequencing methods, platforms, and other difficult-to-standardize conditions result in technical variations in scRNA-seq measurements called batch effects. These challenges merit careful consideration of how to normalize and analyze the data in a way that permits the study of biological variation without conflation by technical variation.

scRNA-seq is often used for cellular identification. The most frequently used workflow involves applying dimensional reduction (such as Uniform Manifold Approximation and Projection (UMAP) \cite{mcinnes_umap_2018} or SPRING \cite{weinreb_spring_2018}) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells. These groups are then often manually annotated according to expression levels of known marker genes. Machine learning classifiers such as support vector machines are also commonly used to identify cells after training with labeled data \cite{abdelaal_comparison_2019}. Another common analysis task is trajectory inference, where the goal is to infer developmental trajectories from static time shots (reviewed in \cite{saelens2019comparison,wagner2020lineage}). Trajectory inference arranges cells on a low-dimensional manifold and assigns pseudo-times according to transcriptomic similarity. There are also methods such as partition-based graph abstraction \cite{wolf_paga_2019} that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions. RNA-velocity-based methods like Dynamo \cite{qiu2022mapping} use splicing information or time-resolved data to predict paths of differentiation from RNA data.

Despite the power and utility of these computational tools, most commonly employed methods for analyzing scRNA-seq data share certain limitations. Dimensional reduction is often one of the first steps in visualizing scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships \cite{chari_specious_2021}. Distorted embeddings inconsistently represent distances between individual data points and between cell types, making it difficult to interpret cell fate transitions. Similar issues arise in trajectory inference and RNA-velocity methods because local neighborhoods can also vary substantially with the choice of embedding. Another limitation of these methods is the lack of reproducibility inherent to stochastic machine learning methods like t-SNE and UMAP, which output different results on separate executions even when hyperparameters remain the same \cite{wattenberg_how_2016}. Many algorithms also perform feature selection (e.g., limiting their analysis to the most variable genes or highly-expressed genes) or require choosing parameters such as perplexity, making results even more sensitive to the inclusion or exclusion of different datasets in the analysis.

An ideal analysis algorithm for scRNA-seq data should possess certain desirable properties. First, the algorithm should incorporate explicit assumptions that can be kept consistent between analyses. Second, the algorithm should be deterministic, further enhancing reproducibility. Third, the axes in visualization should be interpretable, intuitive, and remain fixed across analysis and datasets. Fourth, the algorithm should assign a continuous value to cell identity instead of assigning discrete identity labels, thus enabling a more subtle and fine-grained analysis of cellular identity and developmental trajectories that reflect the continuous nature of biology. Finally, the algorithm should be easily able to incorporate datasets from multiple data sources, including the wealth of new single-cell atlases now being generated \cite{quake2022decade}.

Here, we present a new physics-inspired algorithm, single-cell Type Order Parameters (scTOP), that has all of these desired properties. The algorithm is based on the statistical physics concept of an order parameter -- a coarse-grained quantity that can distinguish between phases of matter. Arguably the most famous example of an order parameter is magnetization in a ferromagnetic system (i.e., a system that can spontaneously magnetize). In a ferromagnet, the magnetization measures the alignment of spins and can be used to distinguish magnetic and paramagnetic phases \cite{landau2013statistical,sethna2021statistical}. It is straightforward to generalize the idea of magnetization to more complex settings such as spin glasses and attractor neural networks \cite{amit1985spin}. In the context of epigenetic landscape models of cell identity, this takes the form of generalized magnetizations that measure how aligned a gene expression state is with each of the attractor basins for different cell fates \cite{lang_epigenetic_2014,pusuluri_cellular_2017}. These generalized magnetizations serve as natural order parameters for cellular identity and can be calculated directly from data \cite{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}. The scTOP algorithm builds on previous work by developing methods for integrating scRNA-seq data, greatly expanding the power and utility of this approach.

The inputs to scTOP are the scRNA-seq measurements of query data as well as a reference basis containing the archetypal gene expression profiles for cell types of interest. The outputs of scTOP are the coordinates of the query data in cell type order parameter space (i.e., generalized magnetizations), with each output dimension serving as a measure of similarity between the query data and a distinct cell type in the reference basis. In this way, scTOP naturally projects gene expression deterministically onto a space with dimension equal to the number of cell types in the reference basis. These generalized coordinates can be used for both cell type identification and visualization. One appealing feature of scTOP is that the assumptions of the algorithm are explicit and are contained entirely in the choice of reference basis. This allows direct comparison across datasets and settings since the coordinates of and distances between data points do not change as long as one uses the same reference basis. One important limitation of scTOP worth noting is that it cannot be easily used to identify novel cell fates. The underlying reason for this is that scTOP assumes one already knows what cell types are of interest in the form of a reference basis and, for this reason, is less suited for exploratory work with undocumented cell types.

@@

We have implemented scTOP in an easy-to-use Python package, available on the \href{https://pypi.org/project/scTOP/}{Python Package Index} and \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP}{Github}. The code written for the analyses in this paper is available on \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP-manuscript}{Github}.

\section{Methods}
\begin{figure}
	\centering
		\includegraphics[scale=0.8]{figs/scTOP manuscript fig1 - no aggregate.pdf}
	\caption{Steps involved in the scTOP algorithm. (a) Each cell in the query data is preprocessed independently of the other cells in the dataset. (b) scRNA-seq atlases, such as the Mouse Cell Atlas, are used to define the reference basis in the algorithm. (c) scTOP scores are the projections of query data onto the non-orthogonal subspace of cell types. Since there is no statistical fitting, no tuning parameters are involved. (d) The process of finding scTOP scores is shown in the grey section, and the blue section illustrates cell-type space. This space has as many dimensions as there are cell types in the reference basis. We can visualize this concept in 1, 2, or 3 dimensions, depending on which cell types are most relevant. In the 3-dimensional case, the shadows on the bounds of the scatter plot are included to better illustrate the position of points in 3-dimensional space.}
	\label{FIG:1}
\end{figure}

\subsection{Mathematical Details}
To calculate cell type similarity scores, scTOP uses a linear algebra projection method inspired by the concept of cell types as attractors in a dynamical system \cite{huang_cell_2005}. This projection method has been applied previously to bulk RNA-seq samples, and we will summarize the theoretical background here \cite{lang_epigenetic_2014}. A cell type is a state of the gene regulatory network. A natural way to represent attractors in this system is with an attractor neural network, where attractors correspond to different cell types. If we represent each gene as a node in a network with an associated value measuring gene expression (positive value representing high gene expression, negative value indicating low gene expression), then a cell type may be denoted by a vector of gene expression values $S_i$, where $i=1,\ldots, G$ spans the $G$ genes being measured. The gene expression profiles
corresponding to the $C$ cell type attractors are also $G$ dimensional vectors $\xi_i^\mu$, where $\mu=1,\ldots, C$ spans the $C$ cell types being stored in the network. In what follows, we assume that $G > C$; in other words, the number of genes being measured is greater than the number of cell types.

Cell types are often highly similar in their patterns of gene regulation. Kanter and Sompolinsky \cite{kanter_associative_1987} defined a pattern storage method for spin-glass-like neural networks where even correlated patterns robustly act as attractors. For the same system, they also defined order parameters: generalized magnetizations $a^{\mu}$, where $\mu$ iterates through each of the stored cell type attractors. The $a^\mu$ can be understood as de-correlated versions of the conventional spin glass magnetization order parameter that measures the similarity between a given network state and the network state $\xi^\mu$. 
Explicitly, the $C$ generalized magnetizations $a^\mu$ can be calculated for a gene expression state $S_i$ via the expression
\be \label{op_eqn}
a^\mu= \sum_{\nu=1}^C\sum_{j=1}^G [A_{\mu \nu}]^{-1}\xi_j^\nu S_j.
\ee
where
\be
A_{\mu \nu}= \sum_{j}^G  \xi_j^\mu \xi_j^\nu,
\ee
is the overlap matrix of gene expression profiles of different cell types. Previous works have shown the order parameters $a^{\mu}$ provided an excellent similarity score for analyzing bulk RNA-seq data \cite{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}. However, bulk RNA-seq measurements average gene expression over many cells, rendering it impossible to measure cell fate transitions beyond an average tissue state.

As shown in figure \ref{FIG:1} (a), the first stage of the algorithm is preprocessing and normalization of the input data. In our algorithm, measurements from each cell are normalized independently and do not depend on the expression profiles of any other cells in the dataset. We assume that the scRNA-seq data has been processed, resulting in a gene-wise count matrix: for each cell, each gene has an integer value that counts the number of assigned RNA reads. Genes are then rank-ordered. The rank order is subsequently converted to a z-score representing the percentile rank of the expression of a gene within the cell relative to all other genes being measured (assuming a normal distribution with mean zero and standard deviation one, for the sake of simplicity). For example, a gene that has higher expression than $97.8\%$ of genes being measured is assigned a score of $z=2$, while a gene that has higher expression than half of the genes is assigned a score of $z=0$. The details of the process and its effects on data distributions are described in Appendix section \ref{preprocessing}. The preprocessing step is a key component of the algorithm. By processing each cell independently, the output for one cell is independent of other cells included in the analysis. This is in contrast with other algorithms which normalize gene expression across cells by, for example, selecting the most variable genes.

Constructing a representative, accurate reference basis for the data being analyzed is vital to the scTOP algorithm. This process is shown in figure \ref{FIG:1} (b). First, relevant single-cell RNA-seq atlases are gathered to be processed. For example, in analyzing mouse lung cells, we used the Mouse Cell Atlas \cite{noauthor_mapping_nodate}, which contains single-cell samples of mouse tissues across all major organs. Once the relevant atlases are identified, we take an average across each cell population that corresponds to a distinct cell type. To mitigate the potent effects of noise and avoid training data imbalances, a sufficient number of cells must be included in each population. We have empirically found that the minimum sample size to achieve reasonable results is 100-200 cells. Noise affects the algorithm most strongly when it appears in the basis since the decorrelation operation involved in the projection separates cells according to the canonical expression levels included in the reference basis. After each cell population is averaged to create the gene expression profiles for the desired cell types, each cell type profile is preprocessed separately using the same procedure as that used for processing single cells: normalization followed by z-scoring. This creates a modular reference basis where cell types can easily be replaced, removed, or added at will.

Mathematically, we can always write the sample vector as a sum of the projected components and a component $S^{\perp}_i$  perpendicular to the cell type subspace: 
\be
S_i = \sum_\mu a^{\mu} \xi^{\mu}_i + S^{\perp}_i.
\ee
Biologically, $S^{\perp}$ contains information about biological processes that introduce variation in transcriptional profile such as cell cycle as well as expression resulting from technical and biological noise. We refer to the $a^\mu$ as the scTOP scores. Each component of this vector measures the projection of the cell with gene expression profile $S_i$ onto the $\mu$-the cell type in the reference basis. These scores can be used to accurately classify cell identity and provide a natural visualization of gene expression in the space of possible cell fates (see figure \ref{FIG:1} (e)).

In general, we are interested in calculating scTOP scores for individual cells. However, it is also sometimes useful to compute ``aggregate'' scTOP scores for cellular populations by averaging the mRNA counts \textit{first}, then preprocessing the data. This produces a gene expression profile and a set of scTOP aggregate scores for each population. Aggregate scores tend to be higher because averaging over the RNA counts mitigates the well-documented effects of scRNA-seq noise \cite{hicks2018missing, lahnemann_eleven_2020}; the averaged gene expression profile for the populations are not as sparse as individual profiles and thus are more similar to the averaged profiles that form the reference basis.

 In the appendix, we give a detailed discussion of scTOP, including the effects of basis choice, score distributions for correct and incorrect cell types, and other technical details.

scTOP does not have any stochasticity, require tuning of hyper-parameters, or statistical fitting procedures. The assumptions of the algorithm are explicit: that the gene expression profiles included in the reference basis are truly representative of the relevant cell types and that the reference cell types correspond to distinct cellular populations. We have found that constructing a reference basis for a cell type typically requires at least 100 cells. Additionally, if the cell types are not sufficiently different from one another, as in early embryonic development or certain stromal cells where the same cell type is compared between organs, the resulting scTOP scores may be unreliable and exhibit extreme sensitivity to small changes in the choice of reference basis. More information on the limitations of scTOP may be found in the appendix. Generally, outside these very limited edge cases, we have found that scTOP can reliably distinguish even extremely closely related cells (as shown in the example of pulmonary alveolar epithelial type 1 (AT1) and type 2 (AT2) cell lineages below).

Finally, we note that scTOP is extremely computationally efficient. The core of the scTOP is equation \ref{op_eqn}. This requires a single matrix inversion involving the reference basis that can be precomputed. For this reason, the computational complexity of scTOP scales \emph{linearly} with the number of cells. The end result is that scTOP takes milliseconds to run, even for very large datasets. 
\section{Results}

\subsection{Benchmarking and validation: Classification of cell fate}

>>>>

\documentclass[vruler,JEB]{COB}%%%%where COB is the template name

% \usepackage[nomarkers,figuresonly]{endfloat}
\usepackage{hyperref}
% \usepackage{amsmath}
% \usepackage{physics}
\bibliographystyle{plainnat}
% \usepackage[pdftex]{graphicx}
% \usepackage{xcolor}
\hypersetup{
    colorlinks,
    citecolor=blue,
    filecolor=black,
    linkcolor=red,
    urlcolor=blue
}
\usepackage{xr}
\newcommand{\comment}[1]{{\color{blue}#1}}
\newcommand{\edit}[1]{{\color{purple}#1}}
\newcommand{\del}[1]{{\color{red}\st{#1}}}
\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

\makeatletter
\newcommand*{\addFileDependency}[1]{% argument=file name and extension
\typeout{(#1)}% latexmk will find this if $recorder=0
% however, in that case, it will ignore #1 if it is a .aux or 
% .pdf file etc and it exists! If it doesn't exist, it will appear 
% in the list of dependents regardless)
%
% Write the following if you want it to appear in \listfiles 
% --- although not really necessary and latexmk doesn't use this
%
\@addtofilelist{#1}
%
% latexmk will find this message if #1 doesn't exist (yet)
\IfFileExists{#1}{}{\typeout{No file #1.}}
}\makeatother

\newcommand*{\myexternaldocument}[1]{%
\externaldocument{#1}%
\addFileDependency{#1.tex}%
\addFileDependency{#1.aux}%
}

\myexternaldocument{SI}

@@

\def\be{\begin{equation}} 
\def\ee{\end{equation}}
\newcommand \bea {\begin{eqnarray}} 
\newcommand \eea {\end{eqnarray}} 
\newcommand \sign {\hbox{sign}} 
\newcommand{\nn} {\nonumber}
\newcommand{\mbbE} {\mathbb{E}}
\usepackage{widetext-TI}

\newtheorem{theorem}{Theorem}
\newtheorem{condition}{Condition}

%The author can find the documentation of the above style file and any additional
%supporting files if required from "http://www.ctan.org"

% *** Do not adjust lengths that control margins, column widths, etc. ***

\begin{document}

\supertitle{Research Article}

\title{scTOP: physics-inspired order parameters for cellular identification and visualization}

\author{Maria Yampolskaya$^{1}$, Michael J Herriges$^{2, 3}$, Laertis Ikonomou$^{4, 5}$, Darrell N Kotton$^{2, 3}$, and Pankaj Mehta$^{1, 2, 6, 7}$}

\address{\add{1}{Department of Physics, Boston University, Boston, MA 02215, USA}
\add{2}{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\add{3}{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
\add{4}{Department of Oral Biology. University at Buffalo, The State University of New York, Buffalo, NY, USA}
\add{5}{Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University at Buffalo, The State University of New York, Buffalo, NY, USA}
\add{6}{Faculty of Computing and Data Science, Boston University, Boston, MA 02215, USA}
\add{7}{Biological Design Center, Boston University, Boston, MA 02215, USA}}

\corres{(\email{mariay@bu.edu}; \email{pankajm@bu.edu})}

@@

\begin{abstract}
Advances in single-cell RNA-sequencing (scRNA-seq) provide an unprecedented window into cellular identity. The abundance of data requires new theoretical and computational frameworks to analyze the dynamics of differentiation and integrate knowledge from cell atlases. We present single-cell Type Order Parameters (scTOP): a statistical-physics-inspired approach for quantifying cell identity given a reference basis of cell types. scTOP can accurately classify cells, visualize developmental trajectories, and assess the fidelity of engineered cells. Importantly, scTOP does this without feature selection, statistical fitting, or dimensional reduction (e.g., UMAP, PCA, etc.). We illustrate the power of scTOP using human and mouse datasets. By reanalyzing mouse lung data, we characterize a transient hybrid alveolar type 1/alveolar type 2 cell population. Visualizations of lineage tracing hematopoiesis data using scTOP confirm that a single clone can give rise to multiple mature cell types. We assess the transcriptional similarity between endogenous and donor-derived cells in the context of murine pulmonary cell transplantation. Our results suggest physics-inspired order parameters can be an important tool for understanding differentiation and characterizing engineered cells. scTOP is available as an easy-to-use Python package.
\end{abstract}

\keywords{Cell types, Differentiation, Developmental trajectory, Single-cell RNA-sequencing, Computational method}

\section{Summary statement}
A computational method was developed to quantify cell identity and visualize the dynamics of differentiation. Concepts from statistical physics provide meaningful coordinates for comparing single-cell RNA-sequencing data across experiments.

\section{Introduction}
Humans and other animals have many specialized cells that arise through a complex developmental process  \citep{zeng_what_2022}. At the blastocyst/epiblast stages of embroynic development, cells are pluripotent and possess the flexibility to develop into fates from all three germ layers \citep{Wolpert_2019}. However, as embryos mature, cellular identity becomes less plastic and more specified through a complex spatiotemporal process involving signaling and other environmental cues. The questions of how and why cells end up in one fate or another are fundamental to developmental biology \citep{stephenson_intercellular_2012, zhu2020principles, rand2021geometry}. More practically, understanding the process of cell fate specification is crucial for developing directed differentiation protocols for engineering cells \emph{in vitro}, with important clinical applications for human health and disease \citep{rossant_impact_2011, rowe2019induced, basil2020cellular}.

One powerful set of techniques for characterizing cellular phenotype is transcriptome-wide measurement using single-cell RNA sequencing (scRNA-seq). Typical scRNA-seq experiments involve thousands of genes and dozens to millions of cells. For this reason, this data is intrinsically high-dimensional and requires specialized computational tools for analysis and interpretation \citep{zhu2020principles, setty2019characterization, schiebinger2019optimal}. This task is especially difficult because the technical challenges of sequencing RNA from individual cells result in data that is noisy and sparse \citep{hicks2018missing, lahnemann_eleven_2020, argelaguet2021computational}. Genes with relatively low expression are often recorded as not being expressed at all, resulting in zero counts called dropouts. Experiment-to-experiment differences in sequencing methods, platforms, and other difficult-to-standardize conditions result in technical variations in scRNA-seq measurements called batch effects. These challenges merit careful consideration of how to normalize and analyze the data in a way that permits the study of biological variation without conflation by technical variation.

scRNA-seq is often used for cellular identification. The most frequently used workflow involves applying dimensional reduction (such as Uniform Manifold Approximation and Projection (UMAP) \citep{mcinnes_umap_2018} or SPRING \citep{weinreb_spring_2018}) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells. These groups are then often manually annotated according to expression levels of known marker genes. Machine learning classifiers such as support vector machines are also commonly used to identify cells after training with labeled data \citep{abdelaal_comparison_2019}. Another common analysis task is trajectory inference, where the goal is to infer developmental trajectories from static time shots (reviewed in \citep{saelens2019comparison,wagner2020lineage}). Trajectory inference arranges cells on a low-dimensional manifold and assigns pseudo-times according to transcriptomic similarity. There are also methods such as partition-based graph abstraction \citep{wolf_paga_2019} that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions. RNA-velocity-based methods like Dynamo \citep{qiu2022mapping} use splicing information or time-resolved data to predict paths of differentiation from RNA data.

Despite the power and utility of these computational tools, most commonly employed methods for analyzing scRNA-seq data share certain limitations. Dimensional reduction is often one of the first steps in visualizing scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships \citep{chari_specious_2021}. Distorted embeddings inconsistently represent distances between individual data points and between cell types, making it difficult to interpret cell fate transitions. Similar issues arise in trajectory inference and RNA-velocity methods because local neighborhoods can also vary substantially with the choice of embedding. Another limitation of these methods is the lack of reproducibility inherent to methods like t-SNE and UMAP, which output different results on separate executions even when hyperparameters remain the same \citep{wattenberg_how_2016}. Many algorithms also perform feature selection (e.g., limiting their analysis to the most variable genes or highly-expressed genes) or require choosing parameters such as perplexity, making results even more sensitive to the inclusion or exclusion of different datasets in the analysis.

An ideal analysis algorithm for scRNA-seq data should possess certain desirable properties. First, the algorithm should incorporate explicit assumptions that can be kept consistent between analyses. Second, the results of the algorithm should be reproducible, so that direct comparison between data sets is possible. Third, the axes in visualization should be interpretable, intuitive, and remain fixed across analysis and datasets. Fourth, the algorithm should assign a continuous value to cell identity instead of assigning discrete identity labels, thus enabling a more subtle and fine-grained analysis of cellular identity and developmental trajectories that reflect the continuous nature of biology. Finally, the algorithm should be easily able to incorporate datasets from multiple data sources, including the wealth of new single-cell atlases now being generated \citep{quake2022decade}.

Here, we present a new physics-inspired algorithm, single-cell Type Order Parameters (scTOP), that has all of these desired properties. The algorithm is based on the statistical physics concept of an order parameter -- a coarse-grained quantity that can distinguish between phases of matter. An order parameter takes on different values depending on the phase of the system. Typically, in a disordered phase, the corresponding order parameter is zero, and in an ordered phase it is nonzero. Arguably the most famous example of an order parameter is magnetization in a ferromagnetic system, which is a collection of spins (in the simplest case, spins can be ``up'' or ``down'') which align with one another. In a ferromagnet, the magnetization measures the alignment of spins and can be used to distinguish magnetic and paramagnetic phases \citep{landau2013statistical,sethna2021statistical}. Below a particular temperature (known as the ``critical temperature''), the spins in the ferromagnet align with one another. In a simple model, the order parameter (magnetization, which measures the average spin) is approximately one if all the spins are up and approximately negative one if all the spins are down. When the temperature exceeds the critical temperature, the system changes from the magnetic phase to the paramagnetic phase, and the spins do not align; instead, they are randomly up or down, and the order parameter is approximately zero.

It is straightforward to generalize the idea of magnetization to more complex settings such as spin glasses and attractor neural networks \citep{amit1985spin}. In epigenetic landscape models of cell identity based on attractor networks, this takes the form of generalized magnetizations that measure how aligned a gene expression state is with each of the attractor basins for different cell fates \citep{lang_epigenetic_2014,pusuluri_cellular_2017}. Whereas magnetization for the ferromagnetic system measures whether the system is in a magnetic or paramagnetic phase, these generalized magnetizations measure which phase a gene regulatory network is in (where the different phases correspond to different cell types). These generalized magnetizations serve as natural order parameters for cellular identity and can be calculated directly from data \citep{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}. The scTOP algorithm builds on previous work by developing methods for integrating scRNA-seq data, greatly expanding the power and utility of this approach. See Supplementary Information section \ref{interpretation} for a further discussion on interpreting score values.

The inputs to scTOP are the scRNA-seq measurements of query data as well as a reference basis containing the archetypal gene expression profiles for cell types of interest. The outputs of scTOP are the coordinates of the query data in cell type order parameter space (i.e., generalized magnetizations), with each output dimension serving as a measure of similarity between the query data and a distinct cell type in the reference basis. In this way, scTOP naturally projects gene expression deterministically onto a space with dimension equal to the number of cell types in the reference basis. These generalized coordinates can be used for both cell type identification and visualization. One appealing feature of scTOP is that the assumptions of the algorithm are explicit and are contained entirely in the choice of reference basis. This allows direct comparison across data sets and settings since the coordinates of and distances between data points do not change as long as one uses the same reference basis. One important limitation of scTOP worth noting is that it cannot be easily used to identify novel cell fates. The underlying reason for this is that scTOP assumes one already knows what cell types are of interest in the form of a reference basis and, for this reason, is less suited for exploratory work with undocumented cell types.

@@

We have implemented scTOP in an easy-to-use Python package, available on the \href{https://pypi.org/project/scTOP/}{Python Package Index} and \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP}{Github}. The code for the analyses in this paper is available in a separate \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP-manuscript}{Github repository}.

\section{Results}
\subsection{Benchmarking and validation: Classification of cell fate}

scTOP scores can be used to reliably and accurately predict cell identity. To show this, we applied the algorithm to several scRNA-seq datasets across species and laboratories. scTOP is organism-agnostic as long as an appropriate reference basis is used. Mice and humans are among the most common subjects of scRNA-seq measurements, and we restrict our analysis in this manuscript to data from these organisms. We make use of single-cell atlases such as the Mouse Cell Atlas \citep{han2018mapping} and the Atlas of Lung Development \citep{negretti_single-cell_2021}.

Table \ref{table:1} lists each of the datasets examined in this paper with their respective scTOP score accuracies. For cases where the training (or ``reference'') data source is the same as the test data source, we set aside a randomly-selected subset of cells for the training set and do not use them in the test data. TopN refers to the percent of cells whose true cell types were in the top N scTOP scores, with a score greater than 0.1. Unspecified refers to the percent of cells for which the highest scTOP score was less than 0.1; this represents a failure of the algorithm to confidently identify the cell. The accuracies across species and tissues are high, and the unspecified percentages are very low, even in datasets exceeding a million cells.

\begin{table*}[h!]
\centering
\begin{tabular}{| c | c | c | c | c | c | c |}
\hline
Test data source & Species & Reference data source & Top1 (percent) & Top3 (percent) & Unspecified (percent) & Total cells \\ 
\hline
Kotton lab & Mouse & Mouse Cell Atlas & 97.19 & 99.77 & 2.69 & 4,805 \\ 
\hline
Brain atlas & Mouse & Self & 86.28 & 96.89 & 1.78 & 1,161,041 \\
\hline
CellBench & Human & Self & 100 & 100 & 0 & 2,822 \\
\hline
Lung atlas & Human & Self & 79.78 & 85.16 & 3.32 & 2,952 \\  
\hline
\hline
Data set & Species & Reference Basis & Top1 (percent) & Top3 (percent) & Unspecified (percent) & Number of cell types \\
\hline
Mouse Cell Atlas & Mouse & Tabula Muris & 79.17 & 87.5 & 0 & 48 \\
\hline

\end{tabular}
\caption{Accuracy scores of included sample types for each of the described data sets.}
\label{table:1}
\end{table*}

scTOP performs comparably to state-of-the-art methods for cell identification. \cite{abdelaal_comparison_2019} compared the performance of automatic cell identification algorithms for various datasets, including the CellBench dataset and an older version of the Allen Mouse Brain atlas. They found that the median F1-score for the CellBench dataset ranged from 0.9 to 1; scTOP has a median F1-score of 1 for the same data. The version of the Allen Mouse Brain atlas they analyzed contained three levels of cell population annotations, with 3, 16, and 92 populations for each annotation level. The median F1-score ranged from 0.64 to 1 for the 16-population case and 0 to 0.98 for the 92-population case. The version of the brain atlas we analyzed contained 40 populations, and scTOP had a median F1-score of 0.88 for this dataset.

\subsubsection{Example 1: Lung lineages}
To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from \cite{herriges_durable_2022} onto the mouse cell atlas. The Herriges et al. dataset measures gene expression levels of six different specialized lung types from healthy mice: AT1, AT2, basal, ciliated, secretory, and neuroendocrine (See figure \ref{herriges} (a)). The reference basis used to find the individual scTOP scores for this data set was derived from the Mouse Cell Atlas. Even though the Herriges et al. dataset and the Mouse Cell Atlas were created in different laboratories using different sequencing methods, the resulting scTOP scores distinctly separate the Herriges et al. dataset into the expected cell types. Furthermore, the individual scTOP scores strongly correlate with the expression of lineage marker sets (figure \ref{herriges} (b)).  Altogether this suggests that scTOP provides a reliable method to annotate scRNA-seq datasets without the need for the expertise required to generate lineage marker gene sets.
